### 2016-04-27 msmfs.tex
Version 0.3, ready for review by Steve and Anna.

\usepackage[sans]{tcc_doc} % SDP style file
\usepackage[english]{babel}
\usepackage{amssymb,amsmath} % for the math symbols
\usepackage{mdwlist}
\usepackage{geometry}
\usepackage{marginnote}
\usepackage{datetime}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{mathtools}
\usepackage{hhtensor}
\usepackage{color}
\usepackage{txfonts}

@@

% Enter here some information needed to fill in the template Title to appear
% on the front pages (will be filled in via the \sdpfrontpage command)
\newcommand{\bigdoctitle}{Implementation of the Rau-Cornwell MSMFS algorithm \xspace}
% Title to go in the "Document Status Sheet" Document number
\newcommand{\docnr}{TCC-SDP-151123-2\xspace}
% Context
\newcommand{\context}{(SDP.PIP)}
% Revision
\newcommand{\revision}{2\xspace}
% Author(s)
\newcommand{\docauthor}{T.J.\ Cornwell\xspace}
% Lead author (goes in the footer)
\newcommand{\leadauthor}{T.J.\ Cornwell\xspace}
% Release date (YYYY-MM-DD)
\newcommand{\docudate}{\today\ \currenttime}
% Document classification
\newcommand{\classification}{Unrestricted}
% Status of the document (draft/final/etc.)
\newcommand{\docstatus}{Draft\xspace}

% The definitions below are used for the signature tables in the preamble of
% the document
\newcommand{\firstsigname}{Name 1 - to be filled in}
\newcommand{\firstsigdesignation}{Designation 1 - to be filled in}
\newcommand{\firstsigaffiliation}{Affiliation 1 - to be filled in}

\newcommand{\secondsigname}{Name 2 - to be filled in}
\newcommand{\secondsigdesignation}{Designation 2 - to be filled in}
\newcommand{\secondsigaffiliation}{Affiliation 2 - to be filled in}

% Table with version numbers
\newcommand{\versiontable}{
  \begin{tabularx}{\textwidth}{|X|X|X|X|}
    \hline
    \bf{Version} & {\bf Date of issue} & {\bf Prepared by} & {\bf Comments}\\
    \hline
  \end{tabularx}
}

@@

% Table with affiliations
\newcommand{\organisationtable}{
  \begin{center}
    \sffamily{\bf ORGANISATION DETAILS}\end{center}
  \begin{table}[htb]
    \centering
    \begin{tabular}[htb]{|l|l|}
      \hline
      Name & Science Data Processor Consortium\\
      \hline
    \end{tabular}
  \end{table}
}

@@

@@

 \begin{center}
 \begin{tabularx}{\textwidth}{|l|X|}
     \hline
     \bf{Reference} & \bf{Reference}\\
     \bf{Number} & \\
     \hline
   {\bf [RD01]} & SKA-TEL-SDP-0000041, F. Malan: iPython Performance Model Description\\
   {\bf [RD02]} & SKA-TEL-SDP-0000027, R. Nijboer: Pipelines Element Subsystem Design\\
   {\bf [RD03]} & SKA-TEL-SDP-0000038, R. Bolton: High Priority Science Objectives: Performance Analysis\\
   {\bf [RD04]} & SKA-TEL-SDP-0000028, A.-J. Boonstra: Ingest Pipeline\\
   {\bf [RD05]} & SKA-TEL-SDP-0000029, S. Salvini: Pipelines: Calibration\\
   {\bf [RD06]} & SKA-TEL-SDP-0000030, A. Scaife: Imaging Pipeline\\
   {\bf [RD07]} & SKA-TEL-SDP-0000031, M. Johnston-Hollitt: Science Analysis Pipeline\\
   {\bf [RD08]} & SKA-TEL-SDP-0000017, S. Wijnholds: Baseline-Dependent Averaging\\
   {\bf [RD09]} & SKA-TEL-SDP-0000058, S. Salvini: Fast Fourier Transforms\\
   {\bf [RD10]} & SKA-TEL-SDP-0000057, C. Skipper: Time and Channel Averaging\\
   {\bf [RD13]} & \url{http://www.astron.nl/casacore/trunk/casacore/doc/notes/229.html}\\
   {\bf [RD15]} & Imager.cc C++ source, CASA source code SVN revision 30821, \url{https://svn.cv.nrao.edu/svn/casa/trunk}\\ 
   {\bf [RD17]} & Parameterized deconvolution for wide-band radio synthesis imaging, Urvashi Rao Venkata, 2010, PhD thesis\\
   {\bf [RD18]} & S. Bhatnagar, T. J. (2008). Correcting direction-dependent gains in the deconvolution of radio interferometric images. A\&A, 419-429\\
   {\bf [RD22]} & Rik Jongerius, S. W. (2014). An End-to-End Computing Model for the Square Kilometre Array. IEEE Computer, volume 47, number 9\\
   {\bf [RD24]} & SDP Element Concept. Paul Alexander, Chris Broekema, Simon Ratcliffe, Rosie Bolton, Bojan Nikolic, 2013, SDP-PROP-DR-001-1 (part of SKA SDP Consortium proposal)\\
   {\bf [RD34]} & Cornwell, T. J. (2008). Multi-Scale CLEAN deconvolution of radio synthesis images. IEEE Journal of Selected Topics in Sig. Proc., 2, 793–801.\\
   {\bf [RD35]} & Conway, J. E., Cornwell, T. J., and Wilkinson, P. N. (1990). Multi-Frequency Synthesis - a New Technique in Radio Interferometric Imaging. Monthly Notices of the Royal Astronomical Society, 246, 490.\\
   {\bf [RD36]} & http://www.aanda.org/component/citedby/?task=crossref\&doi=10.1051/0004-6361/201117104 \\  
   {\bf [RD37]} & Junklewitz, H., Bell, M. R., and Ensslin, T. (2015). A new approach to multifrequency synthesis in radio interferometry. Astronomy \& Astrophysics, 581, A59. http://doi.org/10.1051/0004-6361/201423465 \\
   \hline
 \end{tabularx}
\end{center}

% The actual content goes here
\newpage
\section{Introduction}

@@

\clearpage
\section{Scope of the document}

\clearpage
\section{Background}

The imaging performance of a radio synthesis telescope is related to the uv coverage. The uv coverage is dependent on the observing wavelength so diversity in uv can be obtained by observing over a range of frequencies [RD35]. This Multi-Frequency Synthesis (MFS) only works for radio sources in which broadband emission is present, giving a way to relate measurements at different frequencies. There have been multiple attempts at an MFS algorithm (e.g. [RD35]). MFS is usually most important for sources having extended emission. Hence multi-scale algorithms are favoured.

The Multi-Scale, Multi-Frequency Synthesis (MS-MFS) algorithm was developed by Urvashi [RD17], melding together the concepts from multi-scale clean [RD34] and multi-frequency synthesis e.g. [RD35], with the goal of improving the reconstruction of sky brightness from radio interferometric data. The explicit model used for the sky brightness is a collection of blobs of varying scale sizes and strengths, each with spectral behaviour described by a power law expanded into a Taylor series. It has been used in JVLA, ALMA, and ATCA observations [RD36].

The algorithm is an elaboration of the Multiscale CLEAN. It consists of two parts:
\begin{description}
\item[Major cycle] The image residuals for the current model are calculated by Fourier transform of the images constituting the power law expansion and degridding to obtain the predicted visibility, subtraction from the observed visibility, then gridding, and Fourier transformation to the image plane. There are two ways of performing this cycle depending on whether the Taylor series is calculated in image or uv space. Typically 5 - 10 major cycles are required.
\item[Minor cycle] The minor cycle is a greedy algorithm which identifies the dominant candidate scale and removes that using the appropriate PSF centered on that component. The minor cycle repeats this until convergence. Typically hundreds or thousands of iterations are required.
\end{description}

@@

There is an L1 requirement, $SKA1-SYS\_REQ-2324$ specifying that MSMFS must be used (see Table\ref{tab:L1}).

\begin{table}[htp]
\begin{tabular}{|c|L|c|c|c|}
\hline
ID & Requirement & Status & Parent Requirement & Verification \\
\hline
$SKA1-SYS\_REQ-2324$ &Multi-frequency synthesis imaging. All imaging shall construct and make use of frequency dependent image models over the entire observed bandwidth. & Accepted & Reference [2] & Demonstration \\
\hline 
\end{tabular}	
\caption{Relevant L1 requirement}\label{tab:L1}	
\end{table}

This is reflected in the L2 requirements shown in Table\ref{tab:L2}.
\begin{table}[htp]
\begin{tabular}{|L|L|L|L|L|}
\hline
\multicolumn{2}{|c|}{Function}&\multicolumn{3}{c|}{Satisfies requirement}\\
\hline
Number & 	Name & 	Number	& Name	& Description \\
\hline
F1.9 & Calibrate and Image & $SDP\_REQ-390$	&Multi-frequency synthesis
  imaging	& The SDP shall construct and make
  use of frequency-dependent image models over the entire observed bandwidth. \\
\hline
F1.9 & Calibrate and Image & $SDP\_REQ-676$	& Continuum imaging deconvolution
  scales &	While performing continuum
  imaging the SDP shall deconvolve using a maximum number of 10 (TBC) scales.\\
\hline
F1.9 & Calibrate and Image & $SDP\_REQ-677$	& Continuum imaging deconvolution
  Taylor terms &	While performing continuum
  imaging the SDP shall deconvolve using a maximum number of 5 (TBC) Taylor
  terms.\\
\hline
F1.9 & Calibrate and Image & $SDP\_REQ-678$	& Continuum imaging deconvolution
  by sub-band	& While performing continuum
  imaging the SDP shall deconvolve by sub-band (TBC).\\
\hline 
\end{tabular}
\caption{Relevant L2 requirements}	\label{tab:L2}
\end{table}

@@

Finally we could ask is there a superior algorithm already published? (see e.g. [RD37]).

@@

@@

In the Multi-Scale CLEAN [RD??], the sky is modelled as a set of blobs. Performing this decomposition using a gradient search algorithm is very demanding of resources. Instead, in MSCLEAN, a greedy algorithm obtained by generalising the CLEAN algorithm is used. The greedy algorithm locates the amplitude, scale and location of the blob by a straightforward search in the set of residual images, each smoothed with the appropriate blob. For each minor cycle this produces an amplitude, location, and scale. The smoothed residual images and model images are updated, and iteration proceeds.

@@

The MSMFS algorithm decomposes the image into a set of $N_s$ blobs each with $N_t$ Taylor terms.

@@

%\begin{figure}[htb]
%  \centering
%  \includegraphics[width=\textwidth]{./algorithm.pdf}
%  \caption{The MSMFS algorithm as implemented in CASA [RD??]}
%  \label{fig:algoCASA}
%\end{figure}

@@

When moving the MSMFS algorithm to a distributed (loosely coupled) and parallel (tightly coupled) architecture such as the SDP architecture, there are a large number of important factors that must be considered.
\begin{description}
\item[Optimum partitioning] 
\item[Coupling across partitions] e.g. is the same pixel addressed in two different partitions? If so, is there a satisfactory approach to reconciling the two values?
\item[Synchronisation points] e.g. is a global synchronisation point required so that the deconvolutions can be made consistent? Is there an acceptable algorithm for reconciling separate facets?
\item[Amount of CU memory]
\item[Loading of CU memory backplane]
\item[Access to visibility store]
\end{description}

@@

The SDP pipeline framework [RD:??], allows partitioning across multiple axes. In this case, the natural partitions and typical values are:

\begin{description}
\item[Polarisation] Only Stokes I can be modelled using MSMFS
\item[Frequency] O(500) coarse channels are required to avoid bandwidth smearing
\item[Sub-bands] O(1-3) partitions of the coarse channels are required to ensure that the power law approximation is adequate.
\item[Time] O(43200) ~ O(1s) correlator samples are required to avoid smearing on the longest baselines
\item[Facet] O(100) facets to limit the size of the AW projection kernel
%\item[Snapshot time] O(100s) long gridding/degridding sequences are required to limit smearing due to w-plane rotation
\item[Scales] O(5) scales are typically required to model extended structure
\item[Taylor terms] O(5) terms are required to represent the spectral behaviour
\end{description}

According to the logic of the MSMFS algorithm, Sub-bands can be defined as the limit of the power law approximation for brightness so consequently can always be used as the coarsest partition. For Predict and MakeImage the remaining choices are as shown in Figure \ref{tab:partitions} (most rapid first):

@@

The only choice is between the two orderings of Time and Facet. Since deconvolution is non-linear, we prefer that all times be represented in a single facet rather than the other way around.

Hence according to this level of analysis, the best ordering is [Frequency, Time, Facet, Sub-Band]. Parallelisation over Compute Nodes starts from the left to the right, Distribution over Compute Islands works from the right to the left (see Figure \ref{fig:order}):

@@

\begin{itemize}	
\item Construct the facets with some padding, deconvolve each separately, and reconcile after deconvolution by facet to facet broadcast.
\item Construct the facets without padding, send all facets to one compute island, perform deconvolution, distribute models to facets. 
\item Each CI holds a distinct facet, which do not overlap. Synchronise after peak finding and residual updates. After a peak is found, all islands receive the global peak and update accordingly. Since the limit of coupling is assumed to be the CLEAN patch size, all peaks sufficiently far apart can be processed simultaneously.
\end{itemize}

The first approach will inevitably introduce edge effects in the final image revealing the facet grid, while the second will block processing while deconvolution proceeds on the different sub-band compute nodes. We would then benefit from two different types of nodes:
\begin{description}
\item[Predict and MakeImage] Requires the visibility data but not many scratch images. Stalled during MinorCycle.
\item[MinorCycle] Requires a substantial number of scratch images, but not the visibility data. Stalled during Predict and MakeImage.	
\end{description}

@@

We have a modelling problem that is not going to be wished away. The science demands seamless imaging in multiple axes, whereas computing would like to partition the image cube in those same axes. There is a tendency to assert that one axis or another does not have to be seamless but that has not be conclusively argued. Distributing the minor cycle across Compute Islands may be possible in some cases (e.g. spectral line after continuum estimation and subtraction), but we should also be considering how best to parallelise the Minor Cycle on a single, appropriately specified Compute Island.

@@

The CASA tclean program supports parallel execution using python to run copies on different nodes. The visibility data are partitioned in time across the available processors. Gridding and transforms (i.e. Predict and MakeImage) are done separately and in parallel per partition. Then the Images (and weights) are gathered and then normalised. A serial minor cycle is run, blocking all other processing, and then the model image is scattered to all processes. The Major cycle is done in parallel per partition.

@@

The ASKAPsoft code is mirrored in the SDP Github repository. The ASKAPsoft cimager is an MPI-enabled program. Since it was written as parallel program from the beginning, the structure is easier to follow than the CASA version. Distribution over MPI is via frequency, rather than time as in CASA. Predict and MakeImage are distributed using MPI (via a thin bookkeeping layer). The MinorCycle is performed on the master, and thus blocks until the deconvolution steps have completed.

@@

The second criterion is expensive to compute and usually MAXBASE0 is used. Thus for each scale and for each pixel the pixel value at the reference frequency has to be determined by pre-multipling by the inverse Hessian (in term-term space). This means that the cost of the search is:

@@

\begin{equation}
M_{MSMFS} = (N_T (N_s + 1) + N_s) N_{pix} 
\end{equation}

@@

>>>>

\usepackage[sans]{tcc_doc} % SDP style file
\usepackage[english]{babel}
\usepackage{amssymb,amsmath} % for the math symbols
\usepackage{mdwlist}
\usepackage{geometry}
\usepackage{marginnote}
\usepackage{datetime}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{mathtools}
\usepackage{hhtensor}
\usepackage{color}
\usepackage{txfonts}
\usepackage{enumitem}

@@

% Enter here some information needed to fill in the template Title to appear
% on the front pages (will be filled in via the \sdpfrontpage command)
\newcommand{\bigdoctitle}{Implementation of the Rau-Cornwell MSMFS algorithm \xspace}
% Title to go in the "Document Status Sheet" Document number
\newcommand{\docnr}{TCC-SDP-151123-2\xspace}
% Context
\newcommand{\context}{(SDP.PIP)}
% Revision
\newcommand{\revision}{3\xspace}
% Author(s)
\newcommand{\docauthor}{T.J.\ Cornwell\xspace}
% Lead author (goes in the footer)
\newcommand{\leadauthor}{T.J.\ Cornwell\xspace}
% Release date (YYYY-MM-DD)
\newcommand{\docudate}{\today\ \currenttime}
% Document classification
\newcommand{\classification}{Unrestricted}
% Status of the document (draft/final/etc.)
\newcommand{\docstatus}{Draft\xspace}

% The definitions below are used for the signature tables in the preamble of
% the document
\newcommand{\firstsigname}{T.J. Cornwell}
\newcommand{\firstsigdesignation}{Author}
\newcommand{\firstsigaffiliation}{TCC}

\newcommand{\secondsigname}{T.J. Cornwell}
\newcommand{\secondsigdesignation}{Lead Consultant}
\newcommand{\secondsigaffiliation}{TCC}

% Table with version numbers
\newcommand{\versiontable}{
  \begin{tabularx}{\textwidth}{|X|X|X|X|}
    \hline
    \bf{Version} & {\bf Date of issue} & {\bf Prepared by} & {\bf Comments}\\
    \hline
    0.3 & 27/04/2016 & T.J. Cornwell & Ready for review by Anna and Steve \\
    \hline
  \end{tabularx}
}

@@

% Table with affiliations
\newcommand{\organisationtable}{
  \begin{center}
    \sffamily{\bf ORGANISATION DETAILS}\end{center}
  \begin{table}[htb]
    \centering
    \begin{tabular}[htb]{|l|l|}
      \hline
      Name & Tim Cornwell Consulting\\
      \hline
    \end{tabular}
  \end{table}
}

@@

Parallel versions of both variants of MSMFS exist. These work by distributing in time (CASA) and frequency (ASKAPsoft). Parallelisation of the Minor Cycle is not straightforward and consequently both CASA and ASKAPsoft currently use single-threaded inner loops. This is unlikely to be successful for SKA, and instead the Minor Cycle has to be parallelised inside one CI and perhaps additionally distributed over multiple CI's.

In light of this memo, the following recommendations are made concerning the need for MFS:

\begin{enumerate}
\item Perform simulations to explore the need for MSMFS in SKA. Currently an L1 requirement directs SDP to implement MSMFS. However there is no justification given for this very computationally demanding processing. Indeed the promised excellent UV coverage of the SKA telescopes makes it at least questionable, although arguing from the work of Urvashi [RD??] does indicate that with MFS, the dynamic range would be limited to $10^4$ for wide frequency span. Either CASA or ASKAPsoft could be used for the reconstructions.
\item Adding to the desire to see some more justification of the requirement, we note that the L1 requirements do not address the number of scales required, the number of Taylor terms, or the possibility of sub-banding (all of which appear in the SDP L2 requirements. We note three possible outcomes of the simulations: MFS can be ignored, MFS is required and MSMFS is sufficient, or MFS is needed and MFMFS is not sufficient. In the case of the latter outcome, investigation of other deconvolution techniques (e.g. [RD14]) would be required.
\item Assess the importance of the considerations in Section \ref{sec:other}.
\end{enumerate}

The following recommendations address the resource usage:
\begin{enumerate}[resume]
\item Implement both Image (CASA) and Visibility based (ASKAPsoft) calculation of the Taylor term images in the SDP Performance Model. The main reasons to choose one over the other are to do with the relative speed of gridding and FFTs. When it comes to implementation during construction, the costs of implementing both approaches should be minimal, and unimportant compared to the risk of implementing the wrong choice.
\item Update the SDP Performance Model to reflect the MSMFS memory usage:
\begin{equation}
M_{MSMFS} = (N_T (N_s + 1) + N_s) N_{pix}^2 
\end{equation}
\item Update the SDP Performance Model to reflect the MSMFS Flops:
\begin{equation}
Rflop_{MSMFSminor} = N_{minor} N_{pix}^2 N_s N_T	
\end{equation}
\item Prototype a multi-threaded Minor Cycle on one Compute Island. GPUMultiscaleCLEAN might be a good starting point. Do MS and then MSMFS. This can be done without implementing a major cycle.
\item Model and then, if promising, prototype a distributed Minor Cycle on multiple CI's where synchronisation occurs every iteration. Do MS and then MSMFS. This can be done without implementing a major cycle.\end{enumerate}

The following recommendations address wider issues:
\begin{enumerate}[resume]
\item The SKAO Science Team should be consulted about the impact of seams in the images or spectral cubes. These may occur both spatially (if the facets are deconvolved separately, or in frequency, because of the disjoint use of MSMFS in the different sub-bands. It seems questionable that these can be tolerated.

\item We should note that MSMFS was developed in a serial platform where computation was expensive and memory access cheap. In the SDP context, the processing is distributed and parallel, and computation is cheap and memory access expensive. Although SDP has no remit to develop new algorithms, it could encourage researchers to address this new context.
\end{enumerate}

@@

 \begin{center}
 \begin{tabularx}{\textwidth}{|l|X|}
     \hline
     \bf{Reference} & \bf{Reference}\\
     \bf{Number} & \\
     \hline
   {\bf [RD01]} & SKA-TEL-SDP-0000027, R. Nijboer: Pipelines Element Subsystem Design\\
   {\bf [RD02]} & SKA-TEL-SDP-0000038, R. Bolton: High Priority Science Objectives: Performance Analysis\\
   {\bf [RD03]} & SKA-TEL-SDP-0000029, S. Salvini: Pipelines: Calibration\\
   {\bf [RD04]} & SKA-TEL-SDP-0000030, A. Scaife: Imaging Pipeline\\
   {\bf [RD05]} & \url{http://www.astron.nl/casacore/trunk/casacore/doc/notes/229.html}\\
   {\bf [RD06]} & Imager.cc C++ source, CASA source code SVN revision 30821, \url{https://svn.cv.nrao.edu/svn/casa/trunk}\\ 
   {\bf [RD07]} & Parameterized deconvolution for wide-band radio synthesis imaging, Urvashi Rao Venkata, 2010, PhD thesis\\
   {\bf [RD08]} & S. Bhatnagar, T. J. (2008). Correcting direction-dependent gains in the deconvolution of radio interferometric images. A\&A, 419-429\\
   {\bf [RD09]} & Rik Jongerius, S. W. (2014). An End-to-End Computing Model for the Square Kilometre Array. IEEE Computer, volume 47, number 9\\
   {\bf [RD10]} & SDP Element Concept. Paul Alexander, Chris Broekema, Simon Ratcliffe, Rosie Bolton, Bojan Nikolic, 2013, SDP-PROP-DR-001-1 (part of SKA SDP Consortium proposal)\\
   {\bf [RD11]} & Cornwell, T. J. (2008). Multi-Scale CLEAN deconvolution of radio synthesis images. IEEE Journal of Selected Topics in Sig. Proc., 2, 793–801.\\
   {\bf [RD12]} & Conway, J. E., Cornwell, T. J., and Wilkinson, P. N. (1990). Multi-Frequency Synthesis - a New Technique in Radio Interferometric Imaging. Monthly Notices of the Royal Astronomical Society, 246, 490.\\
   {\bf [RD13]} & http://www.aanda.org/component/citedby/?task=crossref\&doi=10.1051/0004-6361/201117104 \\  
   {\bf [RD14]} & Junklewitz, H., Bell, M. R., and Ensslin, T. (2015). A new approach to multifrequency synthesis in radio interferometry. Astronomy \& Astrophysics, 581, A59. http://doi.org/10.1051/0004-6361/201423465 \\
   {\bf [RD15]} & Cornwell, T.J. (2016) Extension of the SDP performance model for cross data-island connectivity, TCC-SDP-151123-1-1 \\
	{\bf [RD16]} & Rau, U., \& Cornwell, T. J. (2011, June 14). A multi-scale multi-frequency deconvolution algorithm for synthesis imaging in radio interferometry. arXiv.org. EDP Sciences. http://doi.org/10.1051/0004-6361/201117104   \\
	{\bf [RD17]} & Bhatnagar, S., Rau, U., \& Golap, K. (2013). Wide-field wide-band Interferometric Imaging: The WB A-Projection and Hybrid Algorithms. The Astrophysical Journal, 770(2), 91. http://doi.org/10.1088/0004-637X/770/2/91 \\
	\hline
 \end{tabularx}
\end{center}

@@

\section{Scope of the document}

The scope includes only the published Multi-Scale, Multi-Frequency algorithm, existing implementations, and inclusion in the SDP processing pipelines. No other MFS algorithms are explored.

\section{Background}

The imaging performance of a radio synthesis telescope is related to the uv coverage. The uv coverage is dependent on the observing wavelength so diversity in uv can be obtained by observing over a range of frequencies [RD12]. This Multi-Frequency Synthesis (MFS) only works for radio sources in which broadband emission is present, giving a way to relate measurements at different frequencies. There have been multiple attempts at an MFS algorithm (e.g. [RD12]). MFS is usually most important for sources having extended emission. Hence multi-scale algorithms are favoured.

The Multi-Scale, Multi-Frequency Synthesis (MS-MFS) algorithm was developed by Urvashi [RD07], melding together the concepts from multi-scale clean [RD11] and multi-frequency synthesis e.g. [RD12], with the goal of improving the reconstruction of sky brightness from radio interferometric data. The explicit model used for the sky brightness is a collection of blobs of varying scale sizes and strengths, each with spectral behaviour described by a power law expanded into a Taylor series. It has been used in JVLA, ALMA, and ATCA observations [RD13].

The algorithm is an elaboration of the Multiscale CLEAN [RD11]. It consists of two parts:
\begin{description}
\item[Major cycle] The image residuals for the current model are calculated by Fourier transform of the images constituting the power law expansion and degridding to obtain the predicted visibility, subtraction from the observed visibility, then gridding, and Fourier transformation to the image plane. There are two ways of performing this cycle depending on whether the Taylor series is calculated in image or uv space. Typically 5 - 10 major cycles are required.
\item[Minor cycle] The minor cycle is a greedy algorithm which identifies the dominant candidate scale and removes that using the appropriate PSF centered on that component. The minor cycle repeats this until convergence. Typically hundreds or thousands of iterations are required.
\end{description}

@@

There is an L1 requirement [AD03], $SKA1-SYS\_REQ-2324$ specifying that MSMFS must be used for continuum imaging (see Table\ref{tab:L1}).

\begin{table}[htp]
\begin{tabular}{|c|L|c|c|c|}
\hline
ID & Requirement & Status & Parent Requirement & Verification \\
\hline
$SKA1-SYS\_REQ-2324$ &Multi-frequency synthesis imaging. All imaging shall construct and make use of frequency dependent image models over the entire observed bandwidth. & Accepted & Reference & Demonstration \\
\hline 
\end{tabular}	
\caption{Relevant L1 requirement}\label{tab:L1}	
\end{table}

This is reflected in the L2 requirements shown in Table \ref{tab:L2}.
\begin{table}[htp]
\begin{center}
\begin{tabular}{|L|L|L|L|L|}
\hline
\multicolumn{2}{|c|}{Function}&\multicolumn{3}{c|}{Satisfies requirement}\\
\hline
Number & 	Name & 	Number	& Name	& Description \\
\hline
F1.9 & Calibrate and Image & $SDP\_REQ-390$	&Multi-frequency synthesis
  imaging	& The SDP shall construct and make
  use of frequency-dependent image models over the entire observed bandwidth. \\
\hline
F1.9 & Calibrate and Image & $SDP\_REQ-676$	& Continuum imaging deconvolution
  scales &	While performing continuum
  imaging the SDP shall deconvolve using a maximum number of 10 (TBC) scales.\\
\hline
F1.9 & Calibrate and Image & $SDP\_REQ-677$	& Continuum imaging deconvolution
  Taylor terms &	While performing continuum
  imaging the SDP shall deconvolve using a maximum number of 5 (TBC) Taylor
  terms.\\
\hline
F1.9 & Calibrate and Image & $SDP\_REQ-678$	& Continuum imaging deconvolution
  by sub-band	& While performing continuum
  imaging the SDP shall deconvolve by sub-band (TBC).\\
\hline 
\end{tabular}
\caption{Relevant L2 requirements}	\label{tab:L2}
\end{center}
\end{table}

@@

Finally we could ask is there a superior algorithm already published? (see e.g. [RD14]). The degree of tuning required for MSMFS (number of scales and number of Taylor terms) is acceptable in a hands-on data processing scenario but may be not acceptable for a general purpose pipeline.

@@

Urvashi has discussed the use of MSMFS for the JVLA (section 8.4.2.1 of [RD??]. With regard to just ignoring the effect of spectral index variations, she says:

\begin{quotation}

Dynamic-range limits when source spectra are ignored : If continuum imaging is done with only MFS gridding and source spectra are ignored, spectral structure will masquerade as spurious spatial structure. These errors will affect regions of the image both on-source and off-source and their magnitudes depend on the available uv-coverage, the frequency range being covered, the choice of reference frequency, and the intensity and spectral index of the source. A rough rule of thumb for an EVLA-type uv-coverages (see section 6.2.4.2) is that for a point source of with spectral index α = −1.0 measured between 1 and 2 GHz, the peak error obtained if the spectrum is ignored is at a dynamic range of $ < 10^3$. Note that when all sources in the observed region of the sky have similar spectral indices, these errors can be reduced by dividing out an average spectral index (one single number over the entire sky) from the visibilities before imaging the\end{quotation}

If we suppose that the SKA PSF is ten times better than that of the JVLA (an over estimate since the RMS sidelobe level goes only as the inverse of the number of antennas), the dynamic range limitation for SKA1-LOW and SKA1-MID still appears at about $10^4$, which is clearly insufficient. Furthermore, the primary beam will vary across the frequency span, and this will lead to spurious structure if spectral effects are ignored (cf Urvashi). The introduction of sub-bands does ameliorate these problems to some degree but then the continuum structure model is only defined in pieces that are not necessarily consistent at the edges.

@@

In the Multi-Scale CLEAN [RD11], the sky is modelled as a set of blobs. Performing this decomposition using a gradient search algorithm is very demanding of resources. Instead, in MSCLEAN, a greedy algorithm obtained by generalising the CLEAN algorithm is used. The greedy algorithm locates the amplitude, scale and location of the blob by a straightforward search in the set of residual images, each smoothed with the appropriate blob. For each minor cycle this produces an amplitude, location, and scale. The smoothed residual images and model images are updated, and iteration proceeds.

@@

The MSMFS [RD07] algorithm decomposes the image into a set of $N_s$ blobs each with $N_t$ Taylor terms. The following description is drawn from [RD07]

@@

@@

When moving the MSMFS algorithm to a distributed (loosely coupled) and parallel (tightly coupled) architecture such as the SDP architecture, there are a large number of important factors that must be considered.
\begin{description}
\item[Optimum partitioning] 
\item[Coupling across partitions] e.g. is the same pixel addressed in two different partitions? If so, is there a satisfactory approach to reconciling the two values? This is a generic question that should be considered for all distributed and parallel processing.
\item[Synchronisation points] e.g. is a global synchronisation point required so that the deconvolutions can be made consistent? Is there an acceptable algorithm for reconciling separate facets?
\item[Amount of CU memory]
\item[Loading of CU memory backplane] The intrinsic operations in MSMFS are mostly simple SAXBY type operations. Hence the memory backplane will be overloaded, and the CPU data-starved.
\item[Access to visibility store] What access to the visibility store is required? A straightforward implementation of the Major/Minor cycle algorithm loads the observed data every major cycle. However, as discussed in [RD15], this is actually only true of the initialisation and any self-calibration cycles. For other cycles, it is necessary only to load the uvw coordinates and flagging information.
\end{description}

@@

The SDP pipeline framework [RD04], allows partitioning across multiple axes. In this case, the natural partitions and typical values are:

\begin{description}
\item[Polarisation] Only Stokes I can be modelled using MSMFS. Modelling the Faraday Rotation is possible but there is no existing algorithm.
\item[Frequency] O(500) coarse channels are required to avoid bandwidth smearing/ There is no advantage to finer channels (see e.g. section 8.4 in [RD??]).
\item[Sub-bands] O(1-3) partitions of the coarse channels are required to ensure that the power law approximation is adequate.
\item[Time] O(43200) ~ O(1s) correlator samples are required to avoid smearing on the longest baselines
\item[Facet] O(100) facets to limit the size of the AW projection kernel
%\item[Snapshot time] O(100s) long gridding/degridding sequences are required to limit smearing due to w-plane rotation
\item[Scales] O(5) scales are typically required to model extended structure (see section 8.4 in [RD??])
\item[Taylor terms] O(5) terms are required to represent the spectral behaviour for moderately complex spectrum sources (see section 8.4 in [RD??])
\end{description}

According to the logic of the MSMFS algorithm, sub-bands can be defined as the limit of the power law approximation for brightness so consequently can always be used as the coarsest partition. For Predict and MakeImage the remaining choices are as shown in Figure \ref{tab:partitions} (most rapid first):

@@

The choice is between the two orderings of Time and Facet. Since deconvolution is non-linear, we prefer that all times be represented in a single facet rather than the other way around. Hence according to this level of analysis, the best ordering is [Frequency, Time, Facet, Sub-Band]. Parallelisation over Compute Nodes starts from the left to the right, Distribution over Compute Islands works from the right to the left (see Figure \ref{fig:order}):

@@

\begin{itemize}	
\item Construct the facets with some padding, deconvolve each separately, and reconcile after deconvolution by facet to facet broadcast.
\item Construct the facets without padding, send all facets to one compute island, perform deconvolution, distribute models to facets. 
\item Each CI holds a distinct facet, which do not need to overlap. Synchronise in each minor cycle after peak finding. After a peak is found, all islands receive all peaks and choose the same global peak and update accordingly. Since the limit of coupling is assumed to be the CLEAN patch size, all peaks sufficiently far apart can be processed simultaneously.
\end{itemize}

The first approach will inevitably introduce edge effects in the final image revealing the facet grid, while the second will block processing while deconvolution proceeds on the different sub-band compute nodes. We would then benefit from two different types of nodes:
\begin{description}
\item[Predict and MakeImage] Requires the visibility data but not many scratch images. Stalls during MinorCycle.
\item[MinorCycle] Requires a substantial number of scratch images, but not the visibility data. Stalls during Predict and MakeImage.	
\end{description}

@@

However, in the minor cycle, we have a modelling problem that is not going to be wished away. The science demands seamless imaging in multiple axes, whereas computing would like to partition the image cube in those same axes. There is a tendency to assert that one axis or another does not have to be seamless but that has not be conclusively argued. Distributing the minor cycle across Compute Islands may be possible in some cases (e.g. spectral line after continuum estimation and subtraction), but we should also be considering how best to parallelise the Minor Cycle on a single, appropriately specified Compute Island.

@@

The CASA tclean program supports parallel execution using python to run copies on different nodes. This mode of parallel execution fits the CASA execution model. The visibility data are partitioned in time across the available processors. Gridding and transforms (i.e. Predict and MakeImage) are done separately and in parallel per partition. Then the Images (and weights) are gathered and then normalised. A serial minor cycle is run, blocking all other processing, and then the model image is scattered to all processes. The Major cycle is done in parallel per partition.

@@

The ASKAPsoft code is mirrored in the SDP Github repository. The ASKAPsoft cimager is MPI-enabled, written as parallel program from the beginning/ Distribution over MPI is via frequency, rather than time as in CASA. Predict and MakeImage are distributed using MPI (via a thin bookkeeping layer). The MinorCycle is performed on the master, and thus blocks until the deconvolution steps have completed.

@@

The second criterion is expensive to compute and usually MAXBASE0 is used. Thus for each scale and for each pixel the pixel value at the reference frequency has to be determined by pre-multiplying by the inverse Hessian (in term-term space). This means that the cost of the search is:

@@

\begin{equation}
M_{MSMFS} = (N_T (N_s + 1) + N_s) N_{pix}^2 
\end{equation}

@@

\clearpage
\section{Other considerations}
\label{sec:other}

Ignoring the behaviour of the MSMFS algorithm, there are other considerations arising from the use of any MFS technique (this discussion drawn from section 8.4 of [RD07]).
